
\chapter{Related Work}
\label{relatedwork}

There is a wide range of projects providing implemantions for recommender
system. Some of them are described in this chapter to give a quick
overview and comparison.


\section{MyMediaLite}

MyMediaLite~\cite{Gantner2011MyMediaLite} is an open source project developed at the
University of Hildesheim and provides several algorithm for rating
prediction and item prediction. It is written in C\# and can be used with
a command line interface. It also provides a graphical interface to
demonstrate recommender algorithms


\section{PREA (Personalized Recommendation Algorithms Toolkit)}

PREA~\cite{2012arXiv1205.3193L} is an open source project written in Java. It provides
a wide range of recommender algorithms and evaluation metrics to test
them. It is maintained by the Georgia Institute of Technology.


\section{Apache Mahout}

Mahout~\cite{mahout} is an open source library in java. It is implemented
on top of Apache Hadoop, so it uses the map/reduce paradigm. This
means it can run on different independent computers.


\section{Duine Framework}

The Duine Framework~\cite{duine} is an open source project written
in java by the Telematica Instituut/Novay. The recommender of the
Duine Framework combines multiple prediction techniques to exploit
the strengths of the different techniques and to avoid their weaknesses.


\section{Cofi}

Cofi~\cite{cofi} provides an algorithm for the rating prediction
task called Maximum Margin Matrix Factorization. It is open source
and written in C++.


\section{LensKit}

Lenskit~\cite{Ekstrand:2011:RRR:2043932.2043958} is a toolkit which provides several recommender
algorithms and an infrastructure to evaluate them. It is an open source
project by the University of Minnesota


\section{Comparison}
Following a table comparing the algorithms implemented by the different
frameworks.

\begin{table}[h]
\scalebox{0.53}{
\begin{tabular}{c|l|cccccc} \toprule
    Category & Feature & PREA & Mahout & Duine & Cofi & MyMedia & \textbf{recsyslab}\\ \midrule
    \multirow{2}{*}{Baselines} & Constant & O & & & O & O & O\\
                               & User/Item Average & O & O & O &O&O& \\
                               & Random &O&&&&O&O\\\midrule
    \multirow{3}{*}{Memory-based CF}& User-based CF (Su and Khoshgoftaar, 2009)&O&O&O&O&O&O\\
                               & Item-based CF (Sarwar et al., 2001)&O&O&O&O&O&O\\
                                & Default Vote, Inf-User-Frex (Beese et al., 1998)&O&O&&&&\\
                                & Slope-One (Lemire and Maclachlan, 2005)&O&O&&&O&O\\ \midrule
    \multirow{6}{*}{Matrix Factorization}&SVD (Paterek, 2007)&O&O&&O&O&O\\
                                         &NMF(Lee and Seung, 2001)&O&&&&&\\
                                         &PMF(Salakhutdinov and Mnih, 2008b&O&&&&&\\
                                         &Bayesian PMF (Salakhutdinov and Mnih, 2008b)&O&&&&&O\\
                                         &Non-linear PMF (Lawrence and Urtasun, 2009)&O&&&&&\\
                                         &RankMFX (???)&&&&&&O\\ \midrule
    \multirow{2}{*}{Other methods}       &Fast NPCA (Yu et al., 2009&O&&&&&\\
                                         &Rank-based CF (Sun et al., 2011, 2012)&O&&&&&\\ \midrule
    \multirow{5}{*}{Evaluation Metric}   &(N)MAE&O&O&O&O&O&\\
                                         &RMSE&O&O&&O&O&\\
                                         &HLU/NDCG&O&&&&O&\\
                                         &Kendall's Tau, Spearman&O&&&&&\\
                                         &Precision/Recall/F1&&O&&&O&O\\
                                         &ARHR/MRHR&&&&&&O\\ \midrule
    \multirow{6}{*}{Miscelaneous}        &Sparse Vector/Matrix&O&O&O&O&O&O\\
                                         %&Wrapper for other languages&O&&&O&O&\\
                                         &Item Recommender for Positive-only Data&&&&&O&O\\
                                         &Release Year&2011&2005&2009&2004&2009&2013\\
                                         &Language&Java&Java&Java&Java&C\#&Python\\
                                         &License&GPL&LGPL&LGPL&GPL&GPL&GPLv3\\ \bottomrule
\end{tabular}}
\caption{Comparison of different recommender frameworks}
\end{table}
